---
title: "S520 Homework 3"
author: "Jack McShane"
date: '2022-02-19'
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages('ggplot2', repos = "https://cran.us.r-project.org")
library(ggplot2)
```


## Problem 1

- c) Determine the expected value of X. 


\begin{center}
    The expected value for a random variable can be calculated by taking a
    weighted sum where each possible value of the variable is weighted according
    to its probability. This can be done in R as so:
\end{center}

``` {r}
vals <- c(1, 3, 4, 6)
probs <- c(.1, .4, .4, .1)
expectation <- sum(vals*probs)
```

``` {r, echo=FALSE}
print(paste("Expected value: ", expectation))
```
\

- d) Determine the variance of X. 

\begin{center}
    The variance of a random variable measures how much observed values differ
    from the mean or expected value of the variable. It can be calculated using
    R in this manner:
\end{center}

``` {r}
squared_diff <- (vals - expectation)^2
variance <- sum(squared_diff * probs)
```

```{r, echo=FALSE}
print(paste("Variance of X: ", variance))
```
\

- e) Determine the standard deviation of X. 

\begin{center}
    The standard deviation of a random variable's distribution indicates how
    spread values of the random variable are from the mean. It can be calculated
    as so:
\end{center}

``` {r}
stddev <- sqrt(variance)
```

```{r, echo=FALSE}
print(paste("Standard Deviation of X: ", stddev))
```
\


## Problem 2
- a) Determine the pmf of X. 

\begin{center}
    The pmf can be calculated as shown below. The get_probs function calculates
    the probability of each value of X according to the equation given in the
    problem.
\end{center}

``` {r}
get_probs = function(vals) {
  probs <- vector()
  
  for (i in seq_along(vals)){
    
    if (vals[i] == 6){
      probs[i] <- 0
    } else {
      probs[i] <- (7 - vals[i]) / 20
    }
    
  }
  return(probs)
}

vals = c(1, 2, 3, 4, 5, 6)
probs <- get_probs(vals)
```

```{r, echo=FALSE}
print(paste("Probabilities: ", probs))

frame <- data.frame(vals, probs)
ggplot(data = frame, 
          mapping = aes( x = vals, y = probs)) + geom_col(aes(x=vals, y=probs))
```

\

- b) Determine the cdf of X. 
``` {r}
get_cdf = function(pmf_probs) {
  cdf_probs <- vector()
  
  for (i in seq_along(pmf_probs)){
    
    if (i == 1){
      cdf_probs[i] <- pmf_probs[i]
    } else {
      cdf_probs[i] <- pmf_probs[i] + cdf_probs[i - 1]
    }
    
  }
  
  return(cdf_probs)
}

cdf_probs <- get_cdf(probs)
frame <- data.frame(vals, cdf_probs)

ggplot(data = frame,
        mapping = aes(x = vals, y = cdf_probs)) + geom_col(aes(x=vals, y=cdf_probs))

# plot(vals, cdf_probs, type='h')
```
\

- c) Determine the expected value of X. 
``` {r}
expectation <- sum(vals*probs)
print(expectation)
```
\

- d) Determine the variance of X. 
``` {r}
squared_diff <- (vals - expectation)^2
variance <- sum(squared_diff * probs)
print(variance)
```
\

- e) Determine the standard deviation of X. 
``` {r}
stddev <- sqrt(variance)
print(stddev)
```
\


## Problem 3
- a) Determine the pmf of X. 
``` {r}
vals = c(1, 2, 5, 10)
probs <- c((4/10), (1/10), (2/10), (3/10))

print(vals)
print(probs)

plot(vals, probs, type='h')
```
\

- b) Determine the cdf of X. 
``` {r}
cdf_probs = get_cdf(probs)
plot(vals, cdf_probs, type='h')
```
\

- c) Determine the expected value of X. 
``` {r}
expectation <- sum(vals*probs)
print(expectation)
```
\

- d) Determine the variance of X. 
``` {r}
squared_diff <- (vals - expectation)^2
variance <- sum(squared_diff * probs)
print(variance)
```
\

- e) Determine the standard deviation of X. 
``` {r}
stddev <- sqrt(variance)
print(stddev)
```
\


## Problem 11

## Problem 13

## Problem something

## Problem something 2

